A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-10-28 jacobian_calc_CMAME.tex
Start work on Results section

The other goal was to compare the four methods, that is including AD, on the basis of speed. Instead of convergence rate being measured, speed of iteration was measured because it could be so done at the same time as accuracy was being measured given a single tangent-stiffness calculation.  It was assumed that the order that each method was evaluated was unimportant, that evaluating each method sucessively within each solver iteration did not affect their performance individually and that speed of computation did not change with time. These assumptions allowed the test program to run the
same problem with each of the methods at effectively the same time and generate an equal volume of data from each method. It was also assumed that the tangent-stiffness matrix calculation time for each of the methods did not vary based on what values the independent variables held, as they will from iteration to iteration as the Newton solution iterate is updated.  This assumption allowed calculation time measurements to be averaged over all iterations within a solution attempt. The purpose of averaging calculation time measurements was to informally address the extraneous variables of parallel evaluation order and computer system load due to system processes not associated with the study. While not part of the study, the example programs available on the corresponding author's website allow the reader to make a comparison of the methods for themselves on the basis of accuracy and convergence rate for two example nonlinear systems. In the examples, one will notice that the forward-difference method will lose quadratic convergence when the step size selected too large such that accuracy relative to AD is decreased.

Once the selection of what to measure was determined, the task was to find a way to take measurements
that satisfied our goals. The goals of collecting accuracy and speed measurements were achieved by
developing and implementing metrics within the simulation program used in the study. The metric used
to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm of the element-wise
difference between the tangent-stiffness matrix produced by the method being evaluated and the
tangent-stiffness matrix produced by the AD based method, given that both methods were set upon the
same problem. The lower the value of this metric, the closer the other method's Jacobian was to the
AD Jacobian, and therefore the more accurate the method was. The expression for the accuracy metric:

\begin{equation} D = \sqrt{\sum_{i=0}^{n-1}\sum_{j=0}^{n-1} (J_{AD}(i,j) - J_{M}(i,j))^2}
\end{equation}

Where $D$ is distance, n is the number of degrees of freedom, $J_{AD}$ is the Jacobian matrix
produced by the AD based method, M stands for other \emph{M}ethod, and the root of the summed
squared element-wise differences between the matrices is taken.  The metric used to compare the
speed of the methods was the time in seconds taken to resume execution in the calling method after
instructing the implementation of the method being evaluated to run and complete itself. Another
basis of comparison used in the study, but not quite a metric, called computational efficiency is
based on the ratio of calculation time to number of Jacobian matrix elements.

\subsubsection{Materials: Computers, Other Software}

Mesh generation for the test problems run in the study was done using the software package
\emph{Cubit} developed at Sandia National Labs \cite{ref-Cubit}. For the test series in which peridynamic node density was
varied, journal scripts compatible with \emph{Cubit} were used. The rule used to select the
arangement of nodes within the borders designated by simulated material block's dimensions was to
manually choose a node spacing, fit the maximum whole number of nodes in a regular grid within the
block, and then check whether or not this number of nodes matched the number of nodes desired for a
particular test. The same process was used for the multicore series. An example journal script can
be found on the corresponding author's website. For future work, it may be beneficial to automate
and randomize this process for time efficiency and to assist statistical analysis.

The test problems were run the \emph{Stampede} HPC cluster computer housed at \emph{TACC} (Texas
Advanced Computing Center) based out of UT Austin. The single core test series was run using the
normal queue compute nodes, while the multicore test series was run using one to four "large memory"
nodes each having one terrabyte of local RAM. This large amount of memory was required for reasons
explained in \ref{JGAM}.

\subsubsection{Methods: Reduction of Data} The output produced by \emph{Perdigm} including the
additional accuracy and speed measurements it was modified to provide was redirected from the
console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).
Directories and filenames were chosen so that data would be easy to sort by individual test run.
Extraneous information was manually deleted from copied versions of the output text files so that
they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts
averaged accuracy or speed data for a test corresponding to a particular peridynamic node density
over all iterations for that run, and then plotted this averaged data as if it were a function of
peridynamic node density as it varied from test to test in the series.  Similarly, for the multicore
tests, data was averaged and plotted in much the same way, but as a funciton of number of cores used
to solve the problem rather than as a function peridynamic node density which was held constant.
These data reduction and plotting scripts can be found on the corresponding autor's website.

Since samples were not repeated, because there was not resources for it, confidence intervals were
not calculated. However, if results of iterations within the same test problem were considered
samples of the same thing, then a confidence interval could be calculated. This assumption was only made
for the purpose of reporting average values on the plots. But, a cursory review of data showed no 
range overlap among the methods for accuracy or speed data from any test run of either series.

\section{Results and Discussion}

This section contains the averaged results from the study presented in tabular and graphic format.
Table \ref{tab:results} shows the core count, number of nonzero Jacobian matrix elements, load step
average calculation time and load step average accuracy. The units for accuracy are derived from the
expression used to calculate the Jacobian matrix elements and the units used to define bulk material
properties, mentioned in \ref{tcs}.

\begin{table}    
        \caption{Averaged Results, Each Test} \label{tab:results}   
    \begin{tabular}{c c c c }
    Cores & Jacobian Els. & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1.99E6  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{table}

The data presented in \ref{tab:results} will now be shown in graphic form alongside some discussion
of results and trends.

\subsection{Speed Data}
Speed measurements were taken according to the methods mentioned in \ref{tcs}. The averaged results
of these speed measurments for the single core tests can be seen in \ref{fig:serial_speed}.
Average time in seconds to compute a Jacobian matrix element is plotted against number of nonzero
Jacobian matrix ekements.
Complex-step turned out to be the slowest of the four methods, taking the longest average time per
iteration. The next slowest method was central-difference, followed by forward-difference and
automatic-differentiation. Because the x axis of \ref{fig:serial_speed} is on a log scale, it
bears mentioning that while visually it may appear that the trend indicated by the progression of
datapoints on the plot is non-linear, the relationship would be closer to linear. This bodes well
for the scalability of each of the methods by themselves in \emph{Peridigm}.

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_speed}
\end{figure}

For the multicore tests, averaged speed results appear in \ref{fig:multi_speed}. For this series,
rather than varying the number of nonzero Jacobian matrix elements the number of cores used to
solve the problem is varied, while the number of nonzero Jacobian matrix elements remained constant.
Complex-step again was the slowest method, followed by central-difference, automatic-differentiation
and then forward-difference, taking the title of fastest method for this group of tests.
\ref{fig:multi_speed} indicates that when looking at average in-iteration speed, each of the
methods scale linearly with the number of cores used for the given conditions of this group of
tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale
well when additional processors are used. An explanation as to why automatic-differentiation and
forward-differences swapped places in speed ranking relative to the single core results was not
determined.

>>>>

The other goal was to compare the four methods, that is including AD, on the basis of speed. Speed is defined as the total compute-time of one iteration of a Newton step which includes the tangent-stiffness calculation and parallel assembly. The speed of iteration was measured because it could be so done at the same time as accuracy was being measured given a single tangent-stiffness calculation.  It was assumed that the order that each method was evaluated in parallel was unimportant, that evaluating each method sucessively within each solver iteration did not affect their performance individually and that speed of computation did not change with time. These assumptions allowed the test program to run the same problem with each of the methods at effectively the same time and generate an equal volume of data from each method. It was also assumed that the tangent-stiffness matrix calculation time for each of the methods did not vary based on the current configuration, as it changes slightly from iteration to iteration as the Newton solution iterate is updated.  This assumption allowed calculation time measurements to be averaged over all iterations within a single load step. The purpose of averaging calculation time measurements was to informally address the extraneous variables of parallel evaluation order and computer system load due to system processes not associated with the study. While not part of the study, the example programs available on the corresponding author's website allow the reader to make a comparison of the methods for themselves on the basis of accuracy and convergence rate for two example nonlinear systems. In these examples, one will notice that the finite difference methods will lose quadratic convergence when the step size selected too large such that accuracy relative to AD is decreased.

Once the quantities-of-interest (QoIs) for comparison were determined, the task was to find a way to take measurements that satisfied our goals. The goals of collecting accuracy and speed measurements were achieved by developing and implementing metrics within the simulation program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm (analgous to an $l^2$-norm except defined on a matrix) of the element-wise difference between the tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix produced by the AD based method (i.e., the \emph{exact} derivitives), given that both methods were set upon the same problem. The lower the value of this metric, the closer the tangent-stiffness of the method being evaluted (i.e, CS, AD, FD)  was to the AD Jacobian, the more accurate the method was. The expression for the accuracy metric is
%
\begin{equation} 
  D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2}
\end{equation}
%
Where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the AD based method, $M$ is replaced with either CS, FD, or CD depending on the method being compared.  The metric used to compare the speed of the different methods was the time in seconds required to calculate the tangent-stiffness matrix\note[JTF]{Verify the correctness of the this statement.  What was here previously was too wordy}. A final basis of comparison used in the study called \emph{computational efficiency} is based on the ratio of calculation time to the total number of tangent-stiffness matrix elements.

\subsubsection{Computers and other software} 
%
Discritization for the test problems run in the study was done using the software package \emph{Cubit} developed at Sandia National Laboratories \cite{ref-Cubit}. \emph{Cubit} generates a finite element mesh that is internally converted to a ``particle'' discritization internally when calling \emph{Peridigm}, its important to node that the number of computational nodes in \emph{Peridigm} correspond to the number of finite elements in the discritization, not the number of finite element nodes.  An example journal script can be found on the corresponding author's website.

The test problems were run the \emph{Stampede} HPC cluster computer housed at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at Austin. The single core test series was run using the {\tt normal} queue compute nodes, while the parallel test series was run using one to four {\tt large memory} nodes each having one terrabyte of local RAM. This large amount of memory was required for reasons explained in Section~\ref{JGAM}.

\subsubsection{Data Reduction} 
The output produced by \emph{Perdigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  Extraneous information was manually deleted from copied versions of the output text files so that they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem rather than as a function peridynamic node density which was held constant. These data reduction and plotting scripts can be found on the corresponding autor's website.

Since samples were not repeated, because there was not resources for it, confidence intervals were not calculated. However, if results of iterations within the same test problem were considered samples of the same thing, then a confidence interval could be calculated. This assumption was only made for the purpose of reporting average values on the plots. But, a cursory review of data showed no range overlap among the methods for accuracy or speed data from any test run of either series. \note[JTF]{Revise this paragraph.}

\section{Results and Discussion}
%
This section contains the average results from the study presented in tabular and graphic format.  Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) non-zero matrix elements, load step average calculation time and load step average accuracy. The units for accuracy are derived from the
expression used to calculate the TS matrix elements and the units used to define bulk material properties, mentioned in Section~\ref{tcs}.
%
\note[JTF]{Michael, work on cleaning this table up for legibility}
\begin{table}    
        \caption{Averaged Results, Each Test} \label{tab:results}   
    \begin{tabular}{c c c c }
      Cores & TS Elements & Calc. Time (s) & Accuracy Diff. ($\si{\mega\pascal}$)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
    1 & $\num{1.99E6}$  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{table}

\subsection{Speed Data}
Speed measurements were taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these speed measurments for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.  Average time in seconds to compute a TS element is plotted against number of nonzero TS elements.  CS turned out to be the slowest of the four methods, taking the longest average time per iteration. The next slowest method was CD, followed by FD and AD. Because the abcissa of Fig.~\ref{fig:serial_speed} is on a log scale, it bears mentioning that while visually it may appear that the trend indicated by the progression of datapoints on the plot is non-linear, the relationship would be closer to linear \note[JTF]{??? Revise}. This bodes well for the scalability of each of the methods by themselves in \emph{Peridigm}. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in Fig.~\ref{fig:multi_speed}. For this series, rather than varying the number of nonzero TS matrix elements the number of compute cores used to solve the problem is increased, while the number of nonzero TS matrix elements remained constant.  CS again was the slowest method, followed by central-difference, automatic-differentiation and then forward-difference.  There is a somewhat surprisong refersal of FD and AD for these simulations.  Figure~\ref{fig:multi_speed} indicates that when looking at average iteration time, each of the methods scale linearly with the number of cores used for the given conditions of this group of
tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale
well when additional processors are used. An explanation as to why automatic-differentiation and
forward-differences swapped places in speed ranking relative to the single core results was not
determined.
